How to select predictive models for decision-making or causal inference?

### 2023-12-13 how_to_select_causal_models.tex
condense: rm duplicate in intro and rm a sentence in ccl

% maths commands
\newcommand{\indep}{\perp \!\!\! \perp}
\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
\usepackage{mathtools}
\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}

@@

\onecolumn
\maketitle
\begin{abstract}
    \textbf{Objective:}
    As predictive models --\emph{eg} from machine learning-- give likely
    outcomes, they may be used to reason on the effect of an intervention, a
    causal-inference task. The increasing complexity of health data has opened
    the door to a plethora of models, but also the Pandora box of model
    selection: which of these models yield the most valid causal estimates?  In
    this paper, we study model selection procedures in practical settings:
    finite samples settings and without well-specification assumption.
    \textbf{Methods:}
    Elaborate risks build proxies of the causal error using ``nuisance''
    re-weighting to compute it on the observed data. But does computing these
    nuisance adds noise to model selection? Drawing from an extensive empirical
    study based on a simulation and three semi-simulated healthcare datasets
    based on real covariates, we compare different metrics for causal model
    selection.\\
    \textbf{Results:} Among all metrics the classical mean squared error (ie.
    factual $\mu\mathrm{-risk}$) is worse and re-weighting it with propensity score
    ($\mu\mathrm{-risk}_{IPW}$) does not bring much improvements. The
    $R\text{-risk}$, which includes a model of mean outcome and propensity
    scores, leads to the best performances.\\
    \textbf{Conclusions:} We outline a good causal model-selection procedure
    using the so-called $R\text{-risk}$; using flexible estimators to compute
    the nuisance models on the train set; and splitting out 10\% of the data to
    compute risks.
\end{abstract}
\keywords{Model Selection, Treatment Effect, G-formula, Observational Study, Machine Learning, }
\clearpage

@@

Predictive modeling bridges to causal-inference with the so-called \emph{outcome
    models}, or G-computation, G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}. A central challenge of inference of treatment
effects is that of confounding: spurious associations between treatment
allocation and baseline health, \emph{eg} only prescribing a drug to mild cases
\citep{hernan_causal_2020,vanderweele2019principles}.
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Causal inference has been central to epidemiology, typically with methods that
model treatment assignment \cite{austin_moving_2015,casucci2018estimating,grose_use_2020}, based on
propensity scores \cite{rosenbaum_central_1983}. But outcome modeling go
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

% TODO:move below
Standard practices to select models in predictive settings rely on
the error on the outcome
\cite{poldrack2020establishment,varoquaux2022evaluating}. However, as we
will see, these practices may not pick the best models
for causal inference, as they can be misled by inhomogeneities due to
treatment allocation.
%

\section{Objectives and structure of the paper}

@@

We first illustrate the problem of causal model selection and briefly review
prior art. Then we anchor causal model selection in the \emph{potential outcome}
framework and details the causal risks and model-selection procedure. We then
rewrite the so-called $R\text{-risk}$, as a reweighted version of mean squared
difference between the true and estimated individualized treatment effect.
Finally, we conduct a thorough empirical study comparing the different metrics
on a wide variety of data-generative processes based on simulations and real
data, going beyond prior work based on smaller simulation sets
\citep{schuler_comparison_2018,alaa_validating_2019} (Appendix
\ref{apd:prior_work}).

@@

\begin{figure}[b!]
    %\centering
    {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
            causal inference}}

@@

Using a predictor to reason on causal effects relies on contrasting the
prediction of the outcome for a given individual with and without the treatment.
%
Given various predictors of the outcome, which one should we use?
%
Standard predictive modeling or machine-learning practice selects the predictor
that minimizes the expected error. However, this predictor may not be the best
model to reason about causal effects of an intervention as Figure
\ref{fig:toy_example} illustrates. Consider the probability $Y$ of an undesirable outcome
(\emph{eg} death), a binary treatment $A \in \{0, 1\}$, and a covariate $X \in
    \mathbb R$ summarizing the patient health status (eg. the Charlson index
\cite{charlson_new_1987}). We simulate a treatment beneficial (decreases mortality)
for patients with high Charlson scores (bad health status) but with little
effect for patients in good condition (low Charlson scores).

% A natural way to select a predictive model for causal inference would be
% an error measure between a causal quantity such as the CATE and models' estimate. But such error is
% not a ``feasible'' risk: it cannot be computed solely from observed data
% and requires oracle knowledge.

@@

The Neyman-Rubin Potential Outcomes framework
\cite{naimi2023defining,imbens_causal_2015} enables statistical reasoning on
causal treatment effects: Given an outcome $Y \in \mathbb R$ (eg. mortality risk
or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
    \{0, 1\}$ (eg.~a medical procedure), and baseline
covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
distribution, $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However,
we want to model the existence of potential observations (unobserved ie.
counterfactual) that correspond to a different treatment. Thus we want
quantities on the counterfactual distribution $O^{*} = (Y(1), Y(0), X, A) \sim
    \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.

@@

\section{The R-risk, a reweighted oracle metric}\label{theory:r_risk_rewrite}

@@

We generate the response functions using random bases extension, a common method
in biostatistics, \emph{eg} functional regression with splines
\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
vary at specific knots, we control the complexity of the non-linear outcome
models. We use random approximation of Radial Basis Function (RBF) kernels
\cite{rahimi_random_2008} to generate the outcome and treatment functions. RBF
use the same process as polynomial splines but replace polynomial by Gaussian
kernels. Unlike polynomial, Gaussian kernels have decreasing influences in the
input space. This avoids unrealistic divergences of the response surfaces at the
ends of the feature space. We generate 1\,000 datasets based on these functions,
with random overlap parameters. Example shown in Figure
\ref{fig:simulation_examples} and details in \ref{apd:experiments:generation}.

@@

Practically, we recommand a train/test split of 90 \%/10\%. Use most of the data
to estimate both the nuisances and outcome models, then compute the risks on the
remaining test set for model selection (emprical argument in Appendix
\ref{apd:results:data_split}).

@@

\paragraph{Nuisance models: more gain than pain}
%
Predictive models are increasingly used to reason about causal effects, for
instance in precision medicine to drive individualized decision. Our results
highlight that they should be selected, validated, and tuned using different
procedures and error measures than those classically used to assess prediction
(estimating the so-called $\mu\text{-risk}$). Rather, selecting the best outcome
model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
to more valid causal estimates. Estimating the $R\text{-risk}$ requires a more
complex procedure than standard cross-validation used \emph{e.g.}~in machine
learning: it involves fitting nuisance models necessary for model evaluation,
though our results show that these can be learned on the same set of data as the
outcome model evaluated. The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
nuisance models a flexible stacking-based family of estimator suffices for good
model selection. For propensity score, we used the Brier score,
which is minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

@@

\paragraph{Extension to binary outcomes}
While we focused on continuous outcomes, in medicine, the target outcome
is often a categorical variable such as mortality status or diagnosis. In
this case, it may be interesting to focus on other estimands than the
Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
\cite{austin2017estimating}. While the odds ratio is natural for
case-control studies \cite{rothman2008case}, other measures
can reduce heterogeneity \cite{colnet2023risk}.
In the log domain, the ratios are written as a difference, the
framework studied here can directly apply.
In particular, the log odds ratio is estimated by the common
cross-entropy loss (or log loss) as in logistic regression.

@@

Prediction models have flourished because their predictions can be easily
demonstrated and validated on left-out data. But they require more
careful validation for decision making, using a metric accounting for the
putative intervention, the $R\text{-risk}$. Even when treated and
untreated population differ little, as in RCTs, the $R\text{-risk}$
brings a sizeable benefit. To facilitate better model selection, we provide Python
code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
Using the $R\text{-risk}$ does make evaluation more
complicated not only because the procedure is more involved, but also
because each intervention requires a dedicated evaluation. However, such
off-policy evaluation remains much less costly than the recommended good
practice of impact evaluation testing the ability of a prediction model
to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
the model-selection procedure puts no constraints on the models used to
build predictive models: it opens the door to evaluating a wide range of
models, from gradient boosting to convolutional neutral, or language
models.

>>>>

% maths commands
\newcommand{\indep}{\perp \!\!\! \perp}
\newcommand{\eg}{\emph{e.g.}}
\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
\usepackage{mathtools}
\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}

@@

\onecolumn
\maketitle
\begin{abstract}
    \textbf{Objective:}
    As predictive models --\eg from machine learning-- give likely
    outcomes, they may be used to reason on the effect of an intervention, a
    causal-inference task. The increasing complexity of health data has opened
    the door to a plethora of models, but also the Pandora box of model
    selection: which of these models yield the most valid causal estimates?  In
    this paper, we study model selection procedures in practical settings:
    finite samples settings and without well-specification assumption.
    \textbf{Methods:}
    Elaborate risks build proxies of the causal error using ``nuisance''
    re-weighting to compute it on the observed data. But does computing these
    nuisance adds noise to model selection? Drawing from an extensive empirical
    study based on a simulation and three semi-simulated healthcare datasets
    based on real covariates, we compare different metrics for causal model
    selection.\\
    \textbf{Results:} Among all metrics the classical mean squared error (ie.
    factual $\mu\mathrm{-risk}$) is worse and re-weighting it with propensity score
    ($\mu\mathrm{-risk}_{IPW}$) does not bring much improvements. The
    $R\text{-risk}$, which includes a model of mean outcome and propensity
    scores, leads to the best performances.\\
    \textbf{Conclusions:} We outline a good causal model-selection procedure
    using the so-called $R\text{-risk}$; using flexible estimators to compute
    the nuisance models on the train set; and splitting out 10\% of the data to
    compute risks.
\end{abstract}
\keywords{Model Selection, Treatment Effect, G-formula, Observational Study, Machine Learning, }
\clearpage

@@

Predictive modeling bridges to causal-inference with the so-called \emph{outcome
    models}, or G-computation, G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}. A central challenge of inference of treatment
effects is that of confounding: spurious associations between treatment
allocation and baseline health, \eg only prescribing a drug to mild cases
\citep{hernan_causal_2020,vanderweele2019principles}.
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Causal inference has been central to epidemiology, typically with methods that
model treatment assignment \cite{austin_moving_2015,casucci2018estimating,grose_use_2020}, based on
propensity scores \cite{rosenbaum_central_1983}. But outcome modeling go
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

\subsection{Objectives and structure of the paper}

@@

We first illustrate the problem of causal model selection and briefly review
prior art. Then we anchor causal model selection in the \emph{potential outcome}
framework and details the causal risks and model-selection procedure. We then
rewrite the so-called $R\text{-risk}$, as a reweighted version of mean squared
difference between the true and estimated individualized treatment effect.
Finally, we conduct a thorough empirical study comparing the different metrics
on a wide variety of data-generative processes based on simulations and real
data, going beyond prior work based on specific simulation settings
\citep{schuler_comparison_2018,alaa_validating_2019} (Appendix
\ref{apd:prior_work}).

@@

Using a predictor to reason on causal effects relies on contrasting the
prediction of the outcome for a given individual with and without the treatment.
%
Given various predictors of the outcome, which one should we use?
%
Standard predictive modeling or machine-learning practice selects the predictor
that minimizes the expected error on the outcome
\citep{poldrack2020establishment,varoquaux2022evaluating}. However, this
predictor may not be the best model to reason about causal effects of an
intervention as Figure \ref{fig:toy_example} illustrates. Consider the
probability $Y$ of an undesirable outcome (\eg death), a binary treatment $A \in
    \{0, 1\}$, and a covariate $X \in \mathbb R$ summarizing the patient health
status (\eg the Charlson index \cite{charlson_new_1987}). We simulate a
treatment beneficial (decreases mortality) for patients with high Charlson
scores (bad health status) but with little effect for patients in good condition
(low Charlson scores).

% A natural way to select a predictive model for causal inference would be

\begin{figure}[h!]
    %\centering
    {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
            causal inference}}

@@

% an error measure between a causal quantity such as the CATE and models' estimate. But such error is
% not a ``feasible'' risk: it cannot be computed solely from observed data
% and requires oracle knowledge.

@@

The Neyman-Rubin Potential Outcomes framework
\cite{naimi2023defining,imbens_causal_2015} enables statistical reasoning on
causal treatment effects: Given an outcome $Y \in \mathbb R$ (\eg mortality risk
or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
    \{0, 1\}$ (\eg a medical procedure), and baseline
covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
distribution, $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However,
we want to model the existence of potential observations (unobserved ie.
counterfactual) that correspond to a different treatment. Thus we want
quantities on the counterfactual distribution $O^{*} = (Y(1), Y(0), X, A) \sim
    \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.

@@

\subsection{R-risk as reweighted oracle metric}\label{theory:r_risk_rewrite}

@@

We generate the response functions using random bases extension, a common method
in biostatistics, \eg functional regression with splines
\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
vary at specific knots, we control the complexity of the non-linear outcome
models. We use random approximation of Radial Basis Function (RBF) kernels
\cite{rahimi_random_2008} to generate the outcome and treatment functions. RBF
use the same process as polynomial splines but replace polynomial by Gaussian
kernels. Unlike polynomial, Gaussian kernels have decreasing influences in the
input space. This avoids unrealistic divergences of the response surfaces at the
ends of the feature space. We generate 1\,000 datasets based on these functions,
with random overlap parameters. Example shown in Figure
\ref{fig:simulation_examples} and details in \ref{apd:experiments:generation}.

@@

Practically, we recommand a train/test split of 90 \%/10\%. Use most of the data
to estimate both the nuisances and candidates, then compute the risks on the
remaining test set for model selection (empirical argument in Appendix
\ref{apd:results:data_split}).

@@

\paragraph{Nuisance models: more gain than pain}
%
Predictive models are increasingly used to reason about causal effects, for
instance in precision medicine to drive individualized decision. Our results
highlight that they should be selected, validated, and tuned using different
procedures and error measures than those classically used to assess prediction
(estimating the so-called $\mu\text{-risk}$). Rather, selecting the best outcome
model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
to more valid causal estimates. Estimating the $R\text{-risk}$ requires a more
complex procedure than standard cross-validation used \eg in machine
learning: it involves fitting nuisance models necessary for model evaluation,
though our results show that these can be learned on the same set of data as the
outcome model evaluated. The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
nuisance models a flexible stacking-based family of estimator suffices for good
model selection. For propensity score, we used the Brier score,
which is minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

@@

\paragraph{Extension to binary outcomes}
While we focused on continuous outcomes, in medicine, the target outcome
is often a categorical variable such as mortality status or diagnosis. In
this case, it may be interesting to focus on other estimands than the
Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
\cite{austin2017estimating}. While the odds ratio is natural for
case-control studies \cite{rothman2008case}, other measures
can reduce heterogeneity \cite{colnet2023risk}.
In the log domain, the ratios are written as a difference, the
framework studied here can directly apply.
% In particular, the log odds ratio is estimated by the common
% cross-entropy loss (or log loss) as in logistic regression.

@@

Prediction models have flourished because their predictions can be easily
demonstrated and validated on left-out data. But they require more
careful validation for decision making, using a metric accounting for the
putative intervention, the $R\text{-risk}$. Even when treated and
untreated population differ little, as in RCTs, the $R\text{-risk}$
brings a sizeable benefit. To facilitate better model selection, we provide Python
code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
% Using the $R\text{-risk}$ does make evaluation more
% complicated not only because the procedure is more involved, but also
% because each intervention requires a dedicated evaluation. However, such
% off-policy evaluation remains much less costly than the recommended good
% practice of impact evaluation testing the ability of a prediction model
% to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
This model-selection procedure puts no constraints on the models used to
build predictive models: it opens the door to evaluating a wide range of
models, from gradient boosting to convolutional neutral, or language
models.
